Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-03-03 Latex/letter/letter.Rmd
txt update

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analized in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. We did not explore alternative methods for GC-bias correction for broad marks as we believe this task is beyond the scope of our work.

@@

> In the main text, we refer the reader to Section B5 of the Web Appendix. To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$. This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article. Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5.

@@

> Results presented in Table 1 pretain to the full model with all possible combinatorial patterns. This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background among the condition. The point of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates. Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates. We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study.

@@

> We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons. First, methods that tend to call narrow peaks within broad regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC. This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment). For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment. Second, and due to a similar reason, these methods will also exhibit unrealistic advantage in performance in terms of Spearman correlation because broad regions of enrichment are much noisier than punctate regions and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from narrow and punctate windows.

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances were genes were expressed but no enrichment of H3K36me3 was observed. Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> We believe our method is quite comparable to other DPC methods in short differential peak calling. These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 25-27). In terms of LFC and Spearman correlation, our method performed quite comparably to ChIPComp, a widely used method for narrow marks. Examples in Figure 4B, 4D, and 4F clearly shows that our model is correctly calling differential peaks similar to other methods. We did see, however, a slight increase in memory consumption of 5GB when analyzing such marks due to the higher resolutions of 250bp. However, such an increase is still well within the available resourses of most high performance computing clusters and even in some personal laptops nowadays.

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling. This effect was also reported by others (Lun \& Smyth, 2015).

@@

>>>>

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.'. We believe that such a list covers the most used algorithms for DPC in the field of bioinformatics.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. A possible explanation for such a lack of effect of GC-content on the analyzed broad ChIP-seq data is that, in such scenarios, the enrichment of counts has a much lower signal-to-noise ratio with respect to background regions in comparison to short epigenomic marks. In addition, the high resolution of conts computed on non-overlapping windows of 500bp makes the effect of GC-content on enrichment regions indistinguishable from the GC-content effect on background regions. We shown in Web Figure 7 results from our GC-content analysis utizilin the epigenomic mark EZH2. Due to these reasons, we did not explore alternative methods for GC-bias correction for broad marks.

@@

> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, such that

$$Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=1|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{1}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{1}\right), $$

$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\delta_{l}\log f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
$$ Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{3}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{3}\right),$$

> where the notation is precisely defined in Section 3 of the main text. Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by a combination of HMM posterior probabilties and/or mixture model posterior proportions/probabilities. Given the chosen constrained parametrization across all HMM states, we use numerical gradient-based optimization to obtain the parameter estimates. In the equations above, $\mathbf{y}_{..j}$ is the sequence of read counts observed across all conditions and replicates on genomic window $j$. On the other hand, the design matrix $x$ is a a matrix of zeros and ones of dimention $G\times L$ that encodes all differential combinatorial patterns of enrichment $L$ across $G$ conditions (see Figure 1 of the main paper for an example). Further details of the implemented RCEM algorithm are detailed in Web Appendix B4.

@@

> Results presented in Table 1 pertain to the full model that utilizes all possible combinatorial patterns in the differential state. This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background by design in our simulation. The purpose of Table 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates. Together with Web Figures 14 and 15, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates. We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study that directly evaluates the performance of model selection in our method.

@@

> We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons. First, methods that tend to incorrectly call narrow peaks within broad regions of enrichment will exhibit inflated performance in terms of LFC, despite not capturing broad signal regions. This is because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment). See example of these  cases in Figure 3 of the paper (narrow H3K36me3 peak calls from diffReps on the rightmost enrichment region of panel B, narrow H3K37me3 peak calls from ChIPComp on the highlighted area of panel D, and punctate EZH2 calls from DiffBind in on the highlighted area of panel F). For broad marks, interest lies not on these single punctate windows but, instead, in broad regions of enrichment. Similarly, for methods that call broad regions of enrichment, which are much noisier than punctate regions, read count correlations calculated from their differential peaks will be inflated in comparison to correlations calculated from those of narrow and punctate windows. Note in Figure 4C, for example, that RSEG exhibits the highest average Spearman correlation among all methods while incorrectly calling broad peaks on short epigenomic marks. Conversely, algortihms that call punctate peaks within broad regions (Figure 3) still exhibit low/negative Spearman correlation, a metric that does not reflect the poor performance of these methods in broad peaks. Lastly, we note that our method, epigraHMM, provides accurate differential peak calls in both broad and short marks.

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed. Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (Ji et. al, 2013; PNAS). Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.

> NAIM, PLEASE SEE MY REPORT IN THE REPO UNDER ./Supp/Report_RNA_vs_ChIP/ FOR A COMPREHENSIVE DISCUSSION ABOUT HOW USING RNA-SEQ TO DEFINE DE GENES AND THEN BENCHMARKING DIFFERENTIAL H3K36ME3 CHIP-SEQ PEAKS CAN BE PROBLEMATIC. WE DO NOT BENCHMARK DIFFERENTIAL PEAK CALLING METHODS IN THE PAPER USING RNA-SEQ. AS I MENTION IN THE PAPER:  'Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites. Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here.'. I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER). THE ONLY PART OF THE PAPER WHERE WE USE RNA-SEQ IS IN CHROMATIN SEGMENTATION (FIGURE 5).

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component. That is, genes that are expected to be actively transcribed in Helas3 (because they are marker genes) were associated with differential peaks and, in addition, were associated with the differential combinatorial pattern of enrichment for H3K36me3 alone. Note that consensus enrichment is a rare state and not expected to be observed in this analysis. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based mixture proportion from the differential combinatorial pattern associated with enrichment for H3K36me3 alone. Note that our selected model had two mixture components in the differential HMM state (either enrichment of H3K36me3 alone, or enrichment of H3K27me3 and EZH2 together). These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> We believe our method is quite comparable to other DPC methods in short differential peak calling. These results are shown in Figure 4 of our paper as well as in the Web Appendix (Web Figures 26-28). In terms of LFC and Spearman correlation, our method performed quite comparably to ChIPComp, a widely used method for narrow marks. Examples in Figure 4B, 4D, and 4F clearly shows that our model is correctly calling differential peaks similar to other methods. We did see, however, a slight increase in memory consumption of 5GB when analyzing such marks due to the higher resolutions of 250bp. However, such an increase is still well within the available resourses of most high performance computing clusters and even in some personal laptops nowadays.

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance in epigraHMM by accounting for input control effect in differential peak calling (See Web Figure 2 for these results). This effect was also reported by others (Lun \& Smyth, 2015). We did not compare in our paper the performance of other methods with and without accounting for input controls. As stated above, analyses reported in our paper were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). For all other methods (epigraHMM, RSEG, and csaw), input controls were not considered and their results are reported as such.

@@

> NAIM, PLEASE SEE MY COMMENT ABOVE ABOUT THIS GENE EXPRESSION ANALYSIS. I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER). PLEASE, SEE MY REPORT UNDER ./Supp/Report_RNA_vs_ChIP/ AND LET ME KNOW IF ANY OF THOSE RESULTS SHOULD BE INCLUDED HERE TO SUPPORT OUR RESPONSES.
